---
title: "Data Visualization with ggplot2"
author: "Evan L. Ray"
date: "February 02, 2018"
output: ioslides_presentation
---

<style>
  .col2 {
    columns: 2 200px;         /* number of columns and width in pixels*/
    -webkit-columns: 2 200px; /* chrome, safari */
    -moz-columns: 2 200px;    /* firefox */
  }
  .col3 {
    columns: 3 100px;
    -webkit-columns: 3 100px;
    -moz-columns: 3 100px;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Warm Up (with a neighbor)

Here are the first few rows of a data set with information about participants in a randomized controlled trial designed to evaluate a substance abuse treatment program.  What are the observational units and variables?  Are the variables categorical or quantitative?

```{r preprocess_HELPrct, echo=FALSE, warning=FALSE, message=FALSE}
library(mosaicData)
library(dplyr)
library(tidyr)
library(ggplot2)

HELPrct <- HELPrct %>%
  select(
    age,
    homeless,
    substance,
    cesd,
    mcs
  ) %>%
  rename(
  dep_score = cesd,
    mental_score = mcs)
```

```{r head_HELPrct, echo = TRUE}
head(HELPrct)
```

## 5 Main Plots (for this class)

<div class="col2">
1. Bar: Categorical
```{r, fig.height = 1.3, fig.width = 3.5, echo = FALSE}
ggplot(data = HELPrct, mapping = aes(x = substance)) +
  geom_bar()
```

2. Histogram: Quantitative
```{r, fig.height = 1.3, fig.width = 3.5, message=FALSE, echo = FALSE}
ggplot(data = HELPrct, mapping = aes(x = dep_score)) +
  geom_histogram()
```

3. Density: Quantitative
```{r, fig.height = 1.3, fig.width = 3.5, echo = FALSE}
ggplot(data = HELPrct, mapping = aes(x = dep_score)) +
  geom_density()
```

4. Boxplot: 1 Quantitative, 1 Categorical
```{r, fig.height = 1.3, fig.width = 3.5, echo = FALSE}
ggplot(data = HELPrct, mapping = aes(x = substance, y = dep_score)) +
  geom_boxplot()
```

5. Scatterplot: 2 Quantitative
```{r, fig.height = 1.3, fig.width = 3.5, echo = FALSE}
ggplot(data = HELPrct, mapping = aes(x = age, y = dep_score)) +
  geom_point()
```

Also: color
</div>

## The Grammar of Graphics

A statistical graphic is a `mapping` of `data` variables
to `aes()`thetic attributes of `geom_`etric objects

```{r, echo = TRUE, eval = FALSE}
ggplot(data = <data_frame>,
    mapping = aes(<attribute1> = <variable1>, <attribute2> = <variable2>),
  ) +
  geom_<geometry type>()
```

```{r, echo = TRUE, fig.height = 1.5}
ggplot(data = HELPrct,
    mapping = aes(x = dep_score, y = mental_score),
  ) +
  geom_point()
```

## What are the `geom_`etric objects?

<div class="col2">
1. `geom_bar`
```{r, fig.height = 1.3, fig.width = 3.5, echo = FALSE}
ggplot(data = HELPrct, mapping = aes(x = substance)) +
  geom_bar()
```

2. `geom_histogram`
```{r, fig.height = 1.3, fig.width = 3.5, message=FALSE, echo = FALSE}
ggplot(data = HELPrct, mapping = aes(x = dep_score)) +
  geom_histogram()
```

3. `geom_density`
```{r, fig.height = 1.3, fig.width = 3.5, echo = FALSE}
ggplot(data = HELPrct, mapping = aes(x = dep_score)) +
  geom_density()
```

4. `geom_boxplot`
```{r, fig.height = 1.3, fig.width = 3.5, echo = FALSE}
ggplot(data = HELPrct, mapping = aes(x = substance, y = dep_score)) +
  geom_boxplot()
```

5. `geom_point`
```{r, fig.height = 1.3, fig.width = 3.5, echo = FALSE}
ggplot(data = HELPrct, mapping = aes(x = age, y = dep_score)) +
  geom_point()
```


## What are the `aes`thetic properties?

1. `x` for horizontal axis (applicable for all geometry types)
2. `y` for vertical axis (applicable for plot types that have a variable on the vertical axis: boxplot and scatter plot)
3. `fill` (for fill color in geometric objects that have an internal area) or `color` (for color of objects that don't have an internal area)

```{r, echo = TRUE, fig.height = 1.5}
ggplot(data = HELPrct,
    mapping = aes(x = dep_score, y = mental_score, color = substance),
  ) +
  geom_point()
```


## 1 Categorical Variable: Bar Plot
```{r, echo = TRUE, fig.width=8, fig.height = 4}
ggplot(data = HELPrct,
    mapping = aes(x = homeless)
  ) +
  geom_bar()
```

## 2 Categorical Variables: Bar Plot
```{r, echo = TRUE, fig.width=8, fig.height=4}
ggplot(data = HELPrct,
    mapping = aes(x = homeless, fill = substance)
  ) +
  geom_bar()
```

## 2 Categorical Variables: Bar Plot
```{r, echo = TRUE, fig.width=8, fig.height=4}
ggplot(data = HELPrct,
    mapping = aes(x = homeless, fill = substance)
  ) +
  geom_bar(position = position_dodge())
```

## 1 Quantitative Variable: Histograms
```{r, echo = TRUE, fig.width=8, fig.height=3}
ggplot(data = HELPrct,
    mapping = aes(x = dep_score),
  ) +
  geom_histogram()
```

## 1 Quantitative Variable: Histograms
```{r, echo = TRUE, fig.width=8, fig.height=3}
ggplot(data = HELPrct,
    mapping = aes(x = dep_score),
  ) +
  geom_histogram(binwidth = 4)
```

## 1 Quantitative Variable: Density Plots
```{r, echo = TRUE, fig.width=8, fig.height=3}
ggplot(data = HELPrct,
    mapping = aes(x = dep_score),
  ) +
  geom_density()
```

## 2 Quantitative Variables: Scatter Plots
```{r, echo = TRUE, fig.width=8, fig.height=3}
ggplot(data = HELPrct,
    mapping = aes(x = dep_score, y = mental_score)
  ) +
  geom_point()
```

## 2 Quantitative Variables: Scatter Plots
```{r, echo = TRUE, fig.width=8, fig.height=3}
ggplot(data = HELPrct,
    mapping = aes(x = dep_score, y = mental_score, color = substance)
  ) +
  geom_point()
```


## 1 Quantitative, 1 Categorical: Box Plots
```{r, echo = TRUE, fig.width=8, fig.height=3}
ggplot(data = HELPrct,
    mapping = aes(x = substance, y = dep_score)
  ) +
  geom_boxplot()
```

## 1 Quantitative, 1 Categorical: Density
```{r, echo = TRUE, fig.width=8, fig.height=3}
ggplot(data = HELPrct,
    mapping = aes(x = dep_score, color = substance)
  ) +
  geom_density()
```

## Summary

| Variables      | Plot Type    | Geometry       | Aesthetics |
|----------------|--------------|----------------|-------------------|
| 1 Categorical  | Bar Plot     | geom_bar       | x |
| 1 Quantitative | Histogram    | geom_histogram | x |
| 1 Quantitative | Density      | geom_density   | x |
| 2 Categorical  | Bar Plot     | geom_bar       | x, fill |
| 2 Quantitative | Scatter Plot | geom_point     | x, y |
| 1 Categorical and 1 Quantitative | Box Plot | geom_boxplot | x (categorical), y (quantitative) |
| 1 Categorical and 1 Quantitative | Density Plot | geom_density | x (quantitative), color (categorical) |

## Common Error 1

No `+` after call to `ggplot()`:

```{r, echo = TRUE, fig.width=8, fig.height=1.5}
ggplot(data = HELPrct,
    mapping = aes(x = dep_score, y = mental_score)
  )
  geom_point()
```

## Common Error 2

`%>%` instead of `+` after call to `ggplot()`:

```{r, echo = TRUE, eval = FALSE}
ggplot(data = HELPrct,
    mapping = aes(x = dep_score, y = mental_score)
  ) %>%
  geom_point()
```

**Error: Mapping must be created by `aes()` or `aes_()`**

## Common Error 3

Forgot `data = `:

```{r, echo = TRUE, fig.width=8, fig.height=1, eval = FALSE}
ggplot(mapping = aes(x = dep_score, y = mental_score)) +
  geom_point()
```

**Error in FUN(X[[i]], ...) : object 'dep_score' not found**

