---
title: "Simple Random Samples: Sections 7.2, 7.3.1, 7.3.2"
output:
  pdf_document:
    keep_tex: true
header-includes:
   - \usepackage{booktabs}
geometry: margin=1.5cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

### Motivating Example (Section 7.2, Example A)

The population is 393 short-stay hospitals.

Let $c_i$ denote the number of patients discharged from the $i$th hospital during January 1968 (our book is old).

```{r, fig.height = 2}
library(ggplot2) # for making plots

# read in data and look at first few rows
hospitals <- read.csv("http://www.evanlray.com/data/rice/Chapter%207/hospitals.txt")
head(hospitals)

# make a histogram
ggplot(data = hospitals, mapping = aes(x = discharges, y = ..density..)) +
  geom_histogram(boundary = 0, bins = 30)
```

Suppose we want to estimate the average number of patients discharged across this population of all hospitals, based on a sample from that population.

### Population

 * $N$ is the population size ($N$ = 393)

 * $c_1, c_2, \ldots, c_N$ are distinct values in the population ($c_1$ = 57, $c_2$ = 35, ...)
    * For this chapter, we regard these as $N$ **fixed constant** values (hence $c_i$); not random!

#### Population Parameters

 * A **population parameter** is a number describing the population.  Examples:
    * **population mean**: $\mu = \frac{1}{N} \sum_{i = 1}^N c_i$
    * **population total**: $\tau = \sum_{i = 1}^N c_i  = N \mu$
    * **population variance**: $\sigma^2 = \frac{1}{N} \sum_{i = 1}^N (c_i - \mu)^2$
 * More generally, a **parameter** is a constant in a probability model, such as the mean and variance of a Normal distribution.

\newpage

### Samples

 * **Sample**: $n$ observations from the population that we will use to estimate population parameters of interest.

 * **Simple random sample**: A sample obtained from a procedure where every sample of size $n$ has the same probability of being drawn from the population
    * For this chapter, **the random mechanism is in how the sample is selected** from the population of fixed values.

### Statistics and Estimators

 * **Statistic**: A number describing the sample.  Depending on context, this may refer to either:
    * A random variable: each sample will result in a different value of the statistic.
        * Example: $\bar{X} = \sum_{i = 1}^n X_i$ (average number of discharges across hospitals in a sample that has not yet been selected; different samples will result in different realized values).
    * The realized value of this random variable for a particular sample.
        * Example: $\bar{x} = \sum_{i = 1}^n x_i$ (average number of discharges across hospitals in a sample we have taken; this is a number).
 * **Estimator**: A statistic (as a random variable) that is used to estimate a population parameter.
    * The random variable $\bar{X}$ may be used as an estimator for the population mean $\mu$.
 * **Estimate**: The realized value of an estimator for a particular sample.
    * For a particular sample, the realized value $\bar{x}$ may be used as an estimate for the population mean $\mu$.

### Summary in a Picture

\includegraphics{pop-sample-quantitative.pdf}

**Note**: Confusingly, our book labels the observations in the population as $x_1, \ldots, x_N$ and in the sample as $x_1, \ldots, x_n$ -- but the first sample item value, $x_1$, may not be the same as the first population item value $c_1$.
