---
title: "Bayesian Analysis of the Paint Drying Example"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
documentclass: extarticle
geometry: margin=0.6in
---

\def\simiid{\stackrel{{\mbox{\text{\tiny i.i.d.}}}}{\sim}}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
require(ggplot2)
require(scales)
require(dplyr)
require(tidyr)
require(readr)
require(mosaic)
```

 * Data Model: $X_1, \ldots, X_5 \simiid \text{Normal}(\theta, 5^2)$

 * We know that $\theta$ is either 10 (quick dry) or 25 (regular).

 * We don't know which value of theta is correct.  Suppose we adopt a prior with equal probablity for each value of $\theta$:
 
$f_{\Theta}(10) = f_{\Theta}(25) = 0.5$

 * Here's some algebra that says the joint pdf of $X_1, \ldots, X_n$ is proportional to the pdf of $\bar{X}$ (based on the data following a $\text{Normal}(\theta, \sigma^2)$ distribution):
 
\begin{align*}
f_{X_1, \ldots, X_n | \Theta}(x_1, \ldots, x_n | \theta) &= \prod_{i=1}^n f_{X_i | \Theta}(x_i | \theta) \\
 &= \prod_{i = 1}^n \frac{1}{\sqrt{2 \pi} \sigma} \exp\left[ \frac{-1}{2 \sigma^2} (x_i - \theta)^2 \right] \\
 &= \left(\frac{1}{\sqrt{2 \pi} \sigma}\right)^n \exp\left[ \frac{-1}{2 \sigma^2} \sum_{i=1}^n (x_i - \theta)^2 \right] \\
 &= \left(\frac{1}{\sqrt{2 \pi} \sigma}\right)^n \exp\left[ \frac{-1}{2 \sigma^2} \sum_{i=1}^n (x_i - \bar{x} + \bar{x} - \theta)^2 \right] \\
 &= \left(\frac{1}{\sqrt{2 \pi} \sigma}\right)^n \exp\left[ \frac{-1}{2 \sigma^2} \sum_{i=1}^n \left\{ (x_i - \bar{x})^2 + 2(x_i - \bar{x}) (\bar{x} - \theta) + (\bar{x} - \theta)^2 \right\} \right] \\
 &= \left(\frac{1}{\sqrt{2 \pi} \sigma}\right)^n \exp\left[ \frac{-1}{2 \sigma^2} \sum_{i=1}^n (x_i - \bar{x})^2 \right] \exp\left[ \frac{-1}{2 \sigma^2} \sum_{i=1}^n 2(x_i - \bar{x}) (\bar{x} - \theta) \right] \exp\left[ \frac{-1}{2 \sigma^2} \sum_{i=1}^n (\bar{x} - \theta)^2 \right] \\
 &= c \exp\left[ \frac{-1}{2 \sigma^2} 2 (\bar{x} - \theta) (n \bar{x} - n \bar{x}) \right] \exp\left[ \frac{-1}{2 \sigma^2} n (\bar{x} - \theta)^2 \right] \\
 &= c \exp\left[ \frac{-1}{2 \sigma^2 / n} (\bar{x} - \theta)^2 \right] \\
 &= c f_{\bar{X} | \Theta}(\bar{x} | \theta)
\end{align*}

 * Where $\bar{X} \sim \text{Normal}(\theta, \sigma^2/n)$

#### Find the posterior probability that $\Theta = 25$ in terms of the pdf of $\bar{X}$ based on $\theta = 10$ and $\theta = 25$, $f_{\bar{X} | \Theta}(20 | 10)$ and $f_{\bar{X} | \Theta}(20 | 25)$.

\newpage

**Posterior probabilities if the two options are $\theta = 25$ and $\theta = 10$:**

```{r, echo = TRUE}
dnorm(20, mean = 25, sd = 5) / (dnorm(20, mean = 25, sd = 5) + dnorm(20, mean = 10, sd = 5))
dnorm(20, mean = 10, sd = 5) / (dnorm(20, mean = 25, sd = 5) + dnorm(20, mean = 10, sd = 5))
```

\vspace{1cm}

**Posterior probabilities if the two options are $\theta = 25$ and $\theta = 0$:**

```{r, echo = TRUE}
dnorm(20, mean = 25, sd = 5) / (dnorm(20, mean = 25, sd = 5) + dnorm(20, mean = 0, sd = 5))
dnorm(20, mean = 0, sd = 5) / (dnorm(20, mean = 25, sd = 5) + dnorm(20, mean = 0, sd = 5))
```

\vspace{1cm}

**Posterior probabilities if the two options are $\theta = 25$ and $\theta = 20$:**

```{r, echo = TRUE}
dnorm(20, mean = 25, sd = 5) / (dnorm(20, mean = 25, sd = 5) + dnorm(20, mean = 20, sd = 5))
dnorm(20, mean = 20, sd = 5) / (dnorm(20, mean = 25, sd = 5) + dnorm(20, mean = 20, sd = 5))
```
