---
title: "Friend or Foe - Likelihood Ratio Tests and CIs"
output:
  pdf_document:
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set Up/Reminder of Example

Recall the friend or foe example: 16 ten month old babies were shown a sequence of videos where a helpful toy assisted a character in going up a hill, or an unhelpful toy pushed the character down the hill.  In the experiment, $x = 14$ out of 16 babies chose the helpful toy to play with when given a choice.

$X \sim \text{Binomial}(16, \theta)$

## 1. Likelihood Ratio Test

Let's conduct a test of the claim that ten month old babies are not capable of processing the contents of the video; they choose their toy at random.

$H_0: \theta = 0.5$

$H_A: \theta \neq 0.5$

#### (a) Write down the form of the likelihood ratio statistic for this test.  Your answer should only involve the random variable $X$ and numbers, but you do not need to simplify your expression other than any clear cancellations.

\vspace{6cm}

#### (b) Find the observed value of the likelihood ratio statistic.  (You may want to use R to do your calculations.)

\vspace{3cm}

#### (c) Using a large-sample approximation to the sampling distribution of the likelihood ratio statistic, find the p-value for this test.  (You will need to use R to do this, using one of the `pchisq` or `qchisq` functions.)

\newpage

## 2. Approximate Confidence Interval by Inverting the Likelihood Ratio Test

#### (a) Suppose you were conducting a likelihood ratio test of the null hypothesis $H_0: \theta = \theta_0$ with a significance level of $\alpha = 0.05$.  For what values of the likelihood ratio statistic would you reject the null hypothesis?  This will be in terms of a critical value for the test, $w^*$.  What is $w^*$?  (You will need to use R to find this, using one of the `pchisq` or `qchisq` functions.)

```{r, echo = FALSE, eval = FALSE}
exp(-0.5 * qchisq(0.95, df = 1))
```

\vspace{2cm}

#### (b) Illustrate how a confidence interval could be obtained by inverting the likelihood ratio test.

Here is a plot of the likelihood function for $\theta$:

```{r, message=FALSE, fig.height=2.5, echo = FALSE}
library(ggplot2)
library(dplyr)
```

```{r, fig.height=2.5, echo = FALSE}
lik <- function(p_vector) {
  result <- vector("numeric", length(p_vector))
  
  for(i in seq_along(p_vector)) {
    p <- p_vector[i]
    result[i] <- dbinom(x = 14, size = 16, p)
  }
  
  return(result)
}

points <- data.frame(
  p = c(0.1, 0.5, 0.9),
  prob_of_data = c(
    dbinom(x = 14, size = 16, prob = 0.1),
    dbinom(x = 14, size = 16, prob = 0.5),
    dbinom(x = 14, size = 16, prob = 0.9)
  )
)

ggplot(data = data.frame(p = c(0, 1)), mapping = aes(x = p)) +
  stat_function(fun = lik, n = 1001) +
  ylab(expression(italic(L)*"("*theta*"|"*x*"=14)")) +
  xlab(expression(theta)) +
  theme_bw()
```

Here is a plot of the likelihood function for $\theta$, but scaled by dividing by the value of the likelihood function at its maximum:

```{r, fig.height=2.5, echo = FALSE}
lik <- function(p_vector) {
  result <- vector("numeric", length(p_vector))
  
  for(i in seq_along(p_vector)) {
    p <- p_vector[i]
    result[i] <- dbinom(x = 14, size = 16, p) / dbinom(x = 14, size = 16, 14/16)
  }
  
  return(result)
}

points <- data.frame(
  p = c(0.1, 0.5, 0.9),
  prob_of_data = c(
    dbinom(x = 14, size = 16, prob = 0.1),
    dbinom(x = 14, size = 16, prob = 0.5),
    dbinom(x = 14, size = 16, prob = 0.9)
  )
)

ggplot(data = data.frame(p = c(0, 1)), mapping = aes(x = p)) +
  stat_function(fun = lik, n = 1001) +
  scale_y_continuous(breaks = seq(from = 0, to = 1, by = 0.1), minor_breaks = seq(from = 0, to = 1, by = 0.05)) +
  ylab(expression(italic(L)*"("*theta*"|"*x*"=14) / "*italic(L)*"(0.875|"*x*"=14)")) +
  xlab(expression(theta)) +
  theme_bw()
```

 * On one of the two figures above (which one?), draw a line at the critical value $w^*$ from part (a) (approximately - this does not need to be exact).
 * Using that line as a reference, indicate the values $\theta_0$ for which you would reject the null hypothesis that $\theta = \theta_0$.
 * Now indicate the values of $\theta$ that would fall in an approximate 95% confidence interval for $\theta$ obtained by inverting the likelihood ratio test.

#### (c) Find the endpoints of your confidence interval using R.

Doing this exactly would require finding the roots of a degree 16 polynomial in $\theta$.  You can do that if you want, but an approximate approach that's good enough would be to calculate the likelihood ratio statistics at a grid of possible values for $\theta$ between 0 and 1, and then find out which values of $\theta$ are in your interval by comparing to the critical value $w^*$.
