---
title: "Large-Sample Normal Approximations to Posterior"
output:
  pdf_document
---

\newcommand{\simiid}{{\mathrel {\mathop {\sim}\limits _{}^{\rm iid}}\,}}
\newcommand{\simapprox}{{\mathrel {\mathop {\sim}\limits _{}^{\rm approx.}}\,}}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
```

## Introduction

We previously considered Bayesian inference for the proportion of M&M’s that are blue based on samples of size $n = 1$, $n = 10$, $n = 20$, and $n = 541$.

Our model was $X \sim \text{Binomial}(n, \theta)$

We considered a non-informative prior of $\Theta \sim \text{Beta}(1, 1)$.

In that case, the posterior is $\Theta | X = x \sim \text{Beta}(1 + x, 1 + n - x)$.

For large $n$, a normal approximation to the posterior is $\Theta | X = x \simapprox \text{Normal}\left(\frac{x}{n}, \left(\frac{n^2}{x} + \frac{n^2}{n - x}\right)^{-1}\right)$

#### n = 1 (I had x = 1 blue M&M in my sample)

**Can't form the normal approximation:** $n - x = 1 - 1 = 0$, so the approximation to the posterior variance is $\frac{n^2}{x} + \frac{n^2}{n - x} = 1 + \frac{1}{0} = \infty$?

```{r, fig.height=2.25}
x <- 1
n <- 1
a_posterior <- 1 + x
b_posterior <- 1 + n - x

ggplot(data = data.frame(theta = c(0, 1)), mapping = aes(x = theta)) +
  stat_function(fun = dbeta,
    args = list(shape1 = a_posterior, shape2 = b_posterior)) +
  geom_vline(xintercept = x/n, color = "orange", linetype = 2) +
  theme_bw()
```

Posterior mean and 95% posterior credible interval based on the exact Beta posterior:

```{r}
a_posterior/(a_posterior + b_posterior)
qbeta(c(0.025, 0.975), shape1 = a_posterior, shape2 = b_posterior)
```

**Can't get anything out of the normal approximation to the posterior**

\newpage

#### n = 10 (I had x = 3 blue M&Ms in my sample)

```{r, fig.height = 2.5}
x <- 3
n <- 10
a_posterior <- 1 + x
b_posterior <- 1 + n - x

ggplot(data = data.frame(theta = c(0, 1)), mapping = aes(x = theta)) +
  stat_function(fun = dbeta,
    args = list(shape1 = a_posterior, shape2 = b_posterior)) +
  stat_function(fun = dnorm,
    args = list(mean = x/n, sd = sqrt(1/(n^2 / x + n^2 / (n - x)))),
    color = "orange",
    linetype = 2) +
  geom_vline(xintercept = x/n, color = "orange", linetype = 2) +
  theme_bw()
```

Posterior mean and 95% posterior credible interval based on the exact Beta posterior:

```{r}
a_posterior/(a_posterior + b_posterior)
qbeta(c(0.025, 0.975), shape1 = a_posterior, shape2 = b_posterior)
```

Approximate posterior mean and 95% posterior credible interval based on the approximate normal posterior:

```{r}
x/n
qnorm(c(0.025, 0.975), mean = x/n, sd = sqrt(1/(n^2 / x + n^2 / (n - x))))
```

\newpage

#### n = 20 (I had x = 7 blue M&Ms in my sample)

```{r, fig.height = 2.5}
x <- 7
n <- 20
a_posterior <- 1 + x
b_posterior <- 1 + n - x

ggplot(data = data.frame(theta = c(0, 1)), mapping = aes(x = theta)) +
  stat_function(fun = dbeta,
    args = list(shape1 = a_posterior, shape2 = b_posterior)) +
  stat_function(fun = dnorm,
    args = list(mean = x/n, sd = sqrt(1/(n^2 / x + n^2 / (n - x)))),
    color = "orange",
    linetype = 2) +
  geom_vline(xintercept = x/n, color = "orange", linetype = 2) +
  theme_bw()
```

Posterior mean and 95% posterior credible interval based on the exact Beta posterior:

```{r}
a_posterior/(a_posterior + b_posterior)
qbeta(c(0.025, 0.975), shape1 = a_posterior, shape2 = b_posterior)
```

Approximate posterior mean and 95% posterior credible interval based on the approximate normal posterior:

```{r}
x/n
qnorm(c(0.025, 0.975), mean = x/n, sd = sqrt(1/(n^2 / x + n^2 / (n - x))))
```

\newpage

#### Sample of large size (As a class, we had x = 138 blue M&Ms in a sample of size n = 541)

```{r, fig.height = 2.5}
x <- 138
n <-541
a_posterior <- 1 + x
b_posterior <- 1 + n - x

ggplot(data = data.frame(theta = c(0, 1)), mapping = aes(x = theta)) +
  stat_function(fun = dbeta,
    args = list(shape1 = a_posterior, shape2 = b_posterior)) +
  stat_function(fun = dnorm,
    args = list(mean = x/n, sd = sqrt(1/(n^2 / x + n^2 / (n - x)))),
    color = "orange",
    linetype = 2) +
  geom_vline(xintercept = x/n, color = "orange", linetype = 2) +
  theme_bw()
```

Posterior mean and 95% posterior credible interval based on the exact Beta posterior:

```{r}
a_posterior/(a_posterior + b_posterior)
qbeta(c(0.025, 0.975), shape1 = a_posterior, shape2 = b_posterior)
```

Approximate posterior mean and 95% posterior credible interval based on the approximate normal posterior:

```{r}
x/n
qnorm(c(0.025, 0.975), mean = x/n, sd = sqrt(1/(n^2 / x + n^2 / (n - x))))
```
