---
title: "Bootstrap Confidence Intervals"
subtitle: "Part 1b - Bootstrap Percentile Confidence Intervals"
output: pdf_document
---

# Gestation Times

We have data that contains a record of every baby born in December 1998 in the United States. For each baby, we have recordings of several different variables - for today, let's focus on the baby's gestation time in weeks (how many weeks pregnant the mother was when she gave birth).

Since we have data for (almost) every baby born in the United States in December 1998, we are in the rare situation where we have data for a full population!  Let's use this to study what's going on with the bootstrap.

Suppose we are interested in the mean gestation time in the population of births in December 1998, and pretend that we just had data from a sample of size $50$.

\newpage

```{r, cache = TRUE, echo = FALSE, fig.height = 2, warning=FALSE, message = FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(mosaic)


babies <- read_csv("https://mhc-stat140-2017.github.io/data/misc/babies1998/babies_dec_1998.csv")
babies <- filter(babies, !is.na(gestation))

set.seed(256)

#dim(babies)
babies_sample <- babies %>% sample_n(50)

pop_mean <- mean(babies$gestation)
samp_mean <- mean(babies_sample$gestation)

pop_mean
samp_mean

ggplot(data = babies, mapping = aes(x = gestation)) +
  geom_histogram(mapping = aes(y = ..density..)) +
  geom_vline(mapping = aes(xintercept = pop_mean), color = "red") +
  xlim(c(15, 50)) +
  ggtitle("Population: 330717 babies")
```


```{r, cache = TRUE, echo = FALSE, fig.height = 2}
ggplot(data = babies_sample, mapping = aes(x = gestation)) +
  geom_histogram(binwidth = 1, mapping = aes(y = ..density..)) +
  geom_vline(mapping = aes(xintercept = samp_mean), color = "cornflowerblue") +
  xlim(c(15, 50)) +
  ggtitle("Sample of 50 babies")
```

```{r, message = FALSE, warning=FALSE, echo = FALSE, cache = TRUE, fig.height=2}
# the dplyr package contains the sample_n function,
# which we use below to draw the bootstrap samples
library(dplyr)

# how many bootstrap samples to take, and storage space for the results
num_samples <- 10^3
estimates <- data.frame(
  estimate = rep(NA, num_samples)
)

# draw many samples from the observed data and calculate mean of each simulated sample
for(i in seq_len(num_samples)) {
  ## Draw a bootstrap sample of size n with replacement from the observed data
  sampled_obs <- babies %>%
    sample_n(size = 50, replace = FALSE)
  
  ## Calculate mean of bootstrap sample
  estimates$estimate[i] <- mean(sampled_obs$gestation)
}

ggplot(data = estimates, mapping = aes(x = estimate)) +
  geom_histogram(binwidth = 0.1, mapping = aes(y = ..density..)) +
  geom_vline(mapping = aes(xintercept = pop_mean), color = "red") +
  geom_vline(mapping = aes(xintercept = samp_mean), color = "cornflowerblue") +
  xlim(c(15, 50)) +
  ggtitle("Sample means for 1000 samples of 50 babies from population")

# find 2.5th percentile and 97.5th percentile; endpoints of a 95% Bootstrap Percentile Interval
#quantile(bootstrap_estimates$estimate, prob = c(0.025, 0.975))
```


```{r, message = FALSE, warning=FALSE, echo = FALSE, cache = FALSE}
# the dplyr package contains the sample_n function,
# which we use below to draw the bootstrap samples
library(dplyr)

# how many bootstrap samples to take, and storage space for the results
num_bootstrap_samples <- 10^3
bootstrap_estimates <- data.frame(
  estimate = rep(NA, num_bootstrap_samples)
)

# draw many samples from the observed data and calculate mean of each simulated sample
for(i in seq_len(num_bootstrap_samples)) {
  ## Draw a bootstrap sample of size n with replacement from the observed data
  bootstrap_resampled_obs <- babies_sample %>%
    sample_n(size = 50, replace = TRUE)
  
  ## Calculate mean of bootstrap sample
  bootstrap_estimates$estimate[i] <- mean(bootstrap_resampled_obs$gestation)
}

# find 2.5th percentile and 97.5th percentile; endpoints of a 95% Bootstrap Percentile Interval
#quantile(bootstrap_estimates$estimate, prob = c(0.025, 0.975))
```


```{r, echo = FALSE, fig.height = 2, cache = FALSE}
library(ggplot2)
ggplot(data = bootstrap_estimates, mapping = aes(x = estimate)) +
  geom_histogram(binwidth = 0.1, mapping = aes(y = ..density..)) +
  geom_vline(mapping = aes(xintercept = pop_mean), color = "red") +
  geom_vline(mapping = aes(xintercept = samp_mean), color = "cornflowerblue") +
  stat_function(fun = dnorm, args = list(mean = mean(babies_sample$gestation), sd = sd(babies_sample$gestation) / sqrt(50)), color = "purple") +
  geom_vline(mapping = aes(xintercept = mean(babies_sample$gestation) - 1.96*sd(babies_sample$gestation) / sqrt(50)), color = "purple") +
  geom_vline(mapping = aes(xintercept = mean(babies_sample$gestation) + 1.96*sd(babies_sample$gestation) / sqrt(50)), color = "purple") +
  xlim(c(15, 50)) +
  ggtitle("Sample means for 1000 samples of 50 babies from sample (with replacement)")
```
