---
title: "Large-Sample Normal Approximation to the Posterior"
output:
  pdf_document:
    keep_tex: yes
header-includes:
  - \usepackage{multicol}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)
library(ggplot2)
library(gridExtra)
```

\def\simiid{\stackrel{{\mbox{\text{\tiny i.i.d.}}}}{\sim}}

# Poisson Model

 * Observe $X_1, \ldots, X_n$; $X_i$ is the number of seedlings in quadrat number $i$.
 * Data Model: $X_i | \Lambda = \lambda \simiid \text{Poisson}(\lambda)$
 * Suppose we use a Gamma prior for $\Lambda$
    * Example: $\Lambda \sim \text{Gamma}(1, 0.01)$ is fairly non-informative
 * Decomposing the log-posterior pdf into contributions from prior and likelihood

$\log\{f_{\Theta | X_1, \ldots, X_n}\} = \log\{ c \cdot f_{\Theta}(\theta) \cdot \prod_{i=1}^n f_{X_i|\Theta}(x_i | \theta) \} = \log(c) + \log\{f_{\Theta}(\theta)\} + \sum_{i=1}^n \log\{ f_{X_i|\Theta}(x_i | \theta) \}$

## Simulation: $\lambda = 10$

#### n = 1

```{r, message=FALSE, warning=FALSE, echo = FALSE}
set.seed(93365)

plot_xmax <- 25

n <- 1
x_sim <- data.frame(x = rpois(n = n, lambda = 10))
```

```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.height = 6}
p_prior <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  stat_function(fun = dgamma, args = list(shape = 1, rate = 0.01, log = TRUE), n = 1001) +
  ggtitle(expression(paste(log, "{", f(theta), "}")))

calc_ind_lik <- function(lambda, x) {
  result <- seq_along(lambda)
  for(j in seq_along(lambda)) {
    result[j] <- sum(dpois(x = x, lambda = lambda[j], log = TRUE))
  }
  return(result)
}

calc_post_plus_lik <- function(lambda, x) {
  dgamma(lambda, shape = 1, rate = 0.01, log = TRUE) +
    calc_ind_lik(lambda, x)
}

p_ind_lik <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  ggtitle(expression(paste(log, "{f(", x[i], "|", theta, ")}")))
for(i in seq_len(n)) {
  p_ind_lik <- p_ind_lik +
    stat_function(fun = calc_ind_lik, args = list(x = x_sim$x[i]), n = 1001)
}

p_combined_lik <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  stat_function(fun = calc_ind_lik, args = list(x = x_sim$x), n = 1001) +
  ggtitle(expression(paste(log, "{f(", x[1], ", ..., ", x[n], "|", theta,")}")))

p_post <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  stat_function(fun = calc_post_plus_lik, args = list(x = x_sim$x), n = 1001) +
  ggtitle(expression(paste(log, "{f(", theta, "|", x[1], ", ..., ", x[n], ")}")))

grid.arrange(p_prior, p_ind_lik, p_combined_lik, p_post, ncol = 1)
```

\newpage

#### n = 10

```{r, message=FALSE, warning=FALSE, echo = FALSE}
set.seed(93365)

n <- 10
x_sim <- data.frame(x = rpois(n = n, lambda = 10))
```

```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.height = 9.5}
p_prior <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  stat_function(fun = dgamma, args = list(shape = 1, rate = 0.01, log = TRUE), n = 1001) +
  ggtitle(expression(paste(log, "{", f(theta), "}")))

calc_ind_lik <- function(lambda, x) {
  result <- seq_along(lambda)
  for(j in seq_along(lambda)) {
    result[j] <- sum(dpois(x = x, lambda = lambda[j], log = TRUE))
  }
  return(result)
}

calc_post_plus_lik <- function(lambda, x) {
  dgamma(lambda, shape = 1, rate = 0.01, log = TRUE) +
    calc_ind_lik(lambda, x)
}

p_ind_lik <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  ggtitle(expression(paste(log, "{f(", x[i], "|", theta, ")}")))
for(i in seq_len(n)) {
  p_ind_lik <- p_ind_lik +
    stat_function(fun = calc_ind_lik, args = list(x = x_sim$x[i]), n = 1001)
}

p_combined_lik <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  stat_function(fun = calc_ind_lik, args = list(x = x_sim$x), n = 1001) +
  ggtitle(expression(paste(log, "{f(", x[1], ", ..., ", x[n], "|", theta,")}")))

p_post <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  stat_function(fun = calc_post_plus_lik, args = list(x = x_sim$x), n = 1001) +
  ggtitle(expression(paste(log, "{f(", theta, "|", x[1], ", ..., ", x[n], ")}")))

p_post_zoom <- ggplot(data = data.frame(x = c(8.3, 11.0)), mapping = aes(x = x)) +
  stat_function(fun = calc_post_plus_lik, args = list(x = x_sim$x), n = 1001) +
  ggtitle(expression(paste(log, "{f(", theta, "|", x[1], ", ..., ", x[n], ")}, zoomed in to [8.3, 11.0], width 2.7")))

grid.arrange(p_prior, p_ind_lik, p_combined_lik, p_post, p_post_zoom, ncol = 1)
```
\newpage

#### n = 100

```{r, message=FALSE, warning=FALSE, echo = FALSE}
set.seed(93365)

n <- 100
x_sim <- data.frame(x = rpois(n = n, lambda = 10))
```

```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.height = 9.5}
p_prior <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  stat_function(fun = dgamma, args = list(shape = 1, rate = 0.01, log = TRUE), n = 1001) +
  ggtitle(expression(paste(log, "{", f(theta), "}")))

calc_ind_lik <- function(lambda, x) {
  result <- seq_along(lambda)
  for(j in seq_along(lambda)) {
    result[j] <- sum(dpois(x = x, lambda = lambda[j], log = TRUE))
  }
  return(result)
}

calc_post_plus_lik <- function(lambda, x) {
  dgamma(lambda, shape = 1, rate = 0.01, log = TRUE) +
    calc_ind_lik(lambda, x)
}

p_ind_lik <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  ggtitle(expression(paste(log, "{f(", x[i], "|", theta, ")}")))
for(i in seq_len(n)) {
  p_ind_lik <- p_ind_lik +
    stat_function(fun = calc_ind_lik, args = list(x = x_sim$x[i]), n = 1001)
}

p_combined_lik <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  stat_function(fun = calc_ind_lik, args = list(x = x_sim$x), n = 1001) +
  ggtitle(expression(paste(log, "{f(", x[1], ", ..., ", x[n], "|", theta,")}")))

p_post <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  stat_function(fun = calc_post_plus_lik, args = list(x = x_sim$x), n = 1001) +
  ggtitle(expression(paste(log, "{f(", theta, "|", x[1], ", ..., ", x[n], ")}")))

p_post_zoom <- ggplot(data = data.frame(x = c(9.4, 10.25)), mapping = aes(x = x)) +
  stat_function(fun = calc_post_plus_lik, args = list(x = x_sim$x), n = 1001) +
  ggtitle(expression(paste(log, "{f(", theta, "|", x[1], ", ..., ", x[n], ")}, zoomed in to [9.4, 10.25], width 0.85")))

grid.arrange(p_prior, p_ind_lik, p_combined_lik, p_post, p_post_zoom, ncol = 1)
```



\newpage

#### n = 1000

```{r, message=FALSE, warning=FALSE, echo = FALSE}
set.seed(93365)

n <- 1000
x_sim <- data.frame(x = rpois(n = n, lambda = 10))
```

```{r, message=FALSE, warning=FALSE, echo = FALSE, fig.height = 9.5}
p_prior <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  stat_function(fun = dgamma, args = list(shape = 1, rate = 0.01, log = TRUE), n = 1001) +
  ggtitle(expression(paste(log, "{", f(theta), "}")))

calc_ind_lik <- function(lambda, x) {
  result <- seq_along(lambda)
  for(j in seq_along(lambda)) {
    result[j] <- sum(dpois(x = x, lambda = lambda[j], log = TRUE))
  }
  return(result)
}

calc_post_plus_lik <- function(lambda, x) {
  dgamma(lambda, shape = 1, rate = 0.01, log = TRUE) +
    calc_ind_lik(lambda, x)
}

p_ind_lik <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  ggtitle(expression(paste(log, "{f(", x[i], "|", theta, ")}")))
for(i in seq_len(n)) {
  p_ind_lik <- p_ind_lik +
    stat_function(fun = calc_ind_lik, args = list(x = x_sim$x[i]), n = 1001)
}

p_combined_lik <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  stat_function(fun = calc_ind_lik, args = list(x = x_sim$x), n = 1001) +
  ggtitle(expression(paste(log, "{f(", x[1], ", ..., ", x[n], "|", theta,")}")))

p_post <- ggplot(data = data.frame(x = c(0, plot_xmax)), mapping = aes(x = x)) +
  stat_function(fun = calc_post_plus_lik, args = list(x = x_sim$x), n = 1001) +
  ggtitle(expression(paste(log, "{f(", theta, "|", x[1], ", ..., ", x[n], ")}")))

p_post_zoom <- ggplot(data = data.frame(x = c(9.95, 10.25)), mapping = aes(x = x)) +
  stat_function(fun = calc_post_plus_lik, args = list(x = x_sim$x), n = 1001) +
  ggtitle(expression(paste(log, "{f(", theta, "|", x[1], ", ..., ", x[n], ")}, zoomed in to [9.95, 10.25], width 0.3")))

grid.arrange(p_prior, p_ind_lik, p_combined_lik, p_post, p_post_zoom, ncol = 1)
```








