---
title: "Example of Maximum Likelihood and Bayesian Inference"
author: "Evan L. Ray"
date: "February 9, 2018"
output: pdf_document
---

# Example: Body Temperatures

* It's generally believed that the average body temperature is 98.6 degrees Farenheit (37 degrees Celsius).
* Let's investigate with measurements of the temperatures of 130 adults.

#### Load libraries

```{r setup, message = FALSE, warning = FALSE}
require(ggplot2)
require(dplyr)
```


#### Read in data set

```{r}
bodytemp = read.table('http://www.amstat.org/publications/jse/datasets/normtemp.dat.txt')
head(bodytemp)
```

#### Set variable `names`

```{r}
names(bodytemp) <- c('temp','sex','hr')
head(bodytemp)
```

#### Set `sex` to a categorical variable (treated as a variable with 2 levels in this data set)

```{r}
bodytemp <- bodytemp %>%
  mutate(
    sex = factor(sex, levels = c("Male","Female"))
  )
```

#### Make a plot

```{r, fig.height=2}
ggplot(data = bodytemp, mapping = aes(x = temp)) +
  geom_density()
```


#### Sample Statistics, 3 ways

Approach 1:

```{r}
mean(bodytemp$temp)
sd(bodytemp$temp)
```

\vspace{0.5cm}

Approach 2:

```{r}
summarize(bodytemp,
  mean_temp = mean(temp),
  sd_temp = sd(temp)
)
```

\vspace{0.5cm}

Approach 3:

```{r}
bodytemp %>% summarize(
  mean_temp = mean(temp),
  sd_temp = sd(temp)
)
```

